---
title: "Generating and visualising Line of Therapy (LOT) cohorts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating and visualising Line of Therapy (LOT) cohorts}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In oncology and hematology (e.g., Multiple Myeloma), patients often receive complex, multi-agent drug regimens sequentially over time. As patients progress or experience toxicity, treatments are adjusted, forming sequential **Lines of Therapy (LOT)** ($1\text{L}, 2\text{L}, 3\text{L}, \dots$).

The `DrugUtilisation` package provides two core functions for Line of Therapy analyses in OMOP CDM databases:

1. **`generateTherapyLineCohortSet()`**: Maps overlapping ingredient exposures into combination regimens and constructs sequential therapy line subcohorts.
2. **`plotTherapyLineSankey()`**: Creates an interactive Plotly Sankey diagram illustrating regimen transitions, tail-regimen aggregation (`"Other Regimens"`), and patient attrition (`"End of Follow-Up"`).

---

# Setting Up Multiple Myeloma Treatment Cohorts

Let's set up a mock OMOP CDM database containing key Multiple Myeloma (MM) treatment ingredients and a base patient cohort.

- **Proteasome Inhibitors**: Bortezomib
- **Immunomodulatory Drugs (IMiDs)**: Lenalidomide
- **Anti-CD38 Monoclonal Antibodies**: Daratumumab

```{r setup, message = FALSE, warning = FALSE}
library(DrugUtilisation)
library(dplyr, warn.conflicts = FALSE)
library(omopgenerics)
library(plotly)

# Create a mock CDM
cdm <- mockDrugUtilisation(numberIndividual = 300, source = "duckdb")

# Extract observation period window to build realistic test exposures
obs <- cdm$observation_period |>
  collect() |>
  filter(observation_period_start_date + 365 < observation_period_end_date) |>
  head(50)

# 1. Create Base Multiple Myeloma Cohort
base_df <- obs |>
  transmute(
    cohort_definition_id = 1L,
    subject_id = person_id,
    cohort_start_date = observation_period_start_date,
    cohort_end_date = observation_period_end_date
  )

cdm <- insertTable(cdm = cdm, name = "mm_base", table = base_df)
cdm$mm_base <- cdm$mm_base |> newCohortTable()

# 2. Create Sequential Treatment Exposures (1L Bortezomib, 2L Lenalidomide, 3L Daratumumab)
exp1 <- obs |>
  transmute(
    cohort_definition_id = 1L,
    subject_id = person_id,
    cohort_start_date = observation_period_start_date + 10,
    cohort_end_date = observation_period_start_date + 80
  )

exp2 <- obs |>
  transmute(
    cohort_definition_id = 2L,
    subject_id = person_id,
    cohort_start_date = observation_period_start_date + 150,
    cohort_end_date = observation_period_start_date + 220
  )

exp3 <- obs |>
  transmute(
    cohort_definition_id = 3L,
    subject_id = person_id,
    cohort_start_date = observation_period_start_date + 280,
    cohort_end_date = observation_period_start_date + 350
  )

df_exp <- bind_rows(exp1, exp2, exp3)

trt_set <- tibble(
  cohort_definition_id = 1:3,
  cohort_name = c("bortezomib", "lenalidomide", "daratumumab")
)

cdm <- insertTable(cdm = cdm, name = "mm_treatments", table = df_exp)
cdm$mm_treatments <- cdm$mm_treatments |> newCohortTable(cohortSetRef = trt_set)

cohortCount(cdm$mm_treatments)
```

---

# Generating Line of Therapy Subcohorts

We call `generateTherapyLineCohortSet()` to map overlapping exposures into combination regimens and construct sequential lines of therapy.

### Key Parameters:

- **`cohort`**: Base patient population cohort table name (`"mm_base"`).
- **`treatmentCohortName`**: Cohort table containing drug exposures (`"mm_treatments"`).
- **`gapEra`**: Treatment interruption threshold (in days, e.g., 60). Gaps longer than `gapEra` trigger a line transition.
- **`maintenanceModalities`**: Specifies single-agent modalities allowed to continue as maintenance without advancing the line number.

```{r generate_lot, message = FALSE, warning = FALSE}
cdm <- generateTherapyLineCohortSet(
  cdm = cdm,
  name = "mm_lot_cohorts",
  cohort = "mm_base",
  treatmentCohortName = "mm_treatments",
  gapEra = 60
)

# Inspect generated LOT cohorts
cdm$mm_lot_cohorts |>
  cohortCount()
```

---

# Visualising Treatment Transitions with Sankey Diagrams

To inspect treatment patterns and transitions across therapy lines, `plotTherapyLineSankey()` generates an interactive Plotly Sankey diagram.

### Key Design Features:

- **Top Regimen Coverage (`coverageThreshold = 0.80`)**: At each line step, regimens representing the top $80\%$ cumulative volume are displayed individually, while long-tail regimens are grouped into `"Other Regimens"`.
- **End of Follow-Up Stream**: Patients who discontinue treatment or exit follow-up transition into `"End of Follow-Up"`.
- **Vertical Column Alignment**: Step positions ($x/y$ coordinates) are strictly enforced so that each therapy line forms a distinct vertical column.

```{r plot_sankey, message = FALSE, warning = FALSE, eval = FALSE}
# Generate interactive Sankey diagram
p <- plotTherapyLineSankey(
  cohort = cdm$mm_lot_cohorts,
  maxLines = 4,
  coverageThreshold = 0.80
)

p
```

---

# Summary

- `generateTherapyLineCohortSet()` automates the construction of complex, multi-agent therapy lines in OMOP CDM datasets.
- `plotTherapyLineSankey()` produces clean, interactive Plotly visualizations suitable for observational research, clinical dashboards, and regulatory submissions.
