---
title: "Generating and visualising Line of Therapy (LOT) cohorts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating and visualising Line of Therapy (LOT) cohorts}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In oncology and hematology (e.g., Multiple Myeloma), patients often receive complex, multi-agent drug regimens sequentially over time. As patients progress or experience toxicity, treatments are adjusted, forming sequential **Lines of Therapy (LOT)** ($1\text{L}, 2\text{L}, 3\text{L}, \dots$).

The `DrugUtilisation` package provides two core functions for Line of Therapy analyses in OMOP CDM databases:

1. **`generateTherapyLineCohortSet()`**: Maps overlapping ingredient exposures into combination regimens and constructs sequential therapy line subcohorts.
2. **`plotTherapyLineSankey()`**: Creates an interactive Plotly Sankey diagram illustrating regimen transitions, tail-regimen aggregation (`"Other Regimens"`), and patient attrition (`"End of Follow-Up"`).

---

# 1. Quick-Start Example (Mock Data)

Let's set up a mock OMOP CDM database containing key Multiple Myeloma (MM) treatment ingredients and a base patient cohort.

- **Proteasome Inhibitors**: Bortezomib
- **Immunomodulatory Drugs (IMiDs)**: Lenalidomide
- **Anti-CD38 Monoclonal Antibodies**: Daratumumab

```{r setup, message = FALSE, warning = FALSE}
library(DrugUtilisation)
library(dplyr, warn.conflicts = FALSE)
library(omopgenerics)
library(plotly)

# Create a mock CDM
cdm <- mockDrugUtilisation(numberIndividual = 300, source = "duckdb")

# Extract observation period window to build realistic test exposures
obs <- cdm$observation_period |>
  collect() |>
  filter(observation_period_start_date + 365 < observation_period_end_date) |>
  head(50)

# 1. Create Base Multiple Myeloma Cohort
base_df <- obs |>
  transmute(
    cohort_definition_id = 1L,
    subject_id = person_id,
    cohort_start_date = observation_period_start_date,
    cohort_end_date = observation_period_end_date
  )

cdm <- insertTable(cdm = cdm, name = "mm_base", table = base_df)
cdm$mm_base <- cdm$mm_base |> newCohortTable()

# 2. Create Sequential Treatment Exposures (1L Bortezomib, 2L Lenalidomide, 3L Daratumumab)
exp1 <- obs |>
  transmute(
    cohort_definition_id = 1L,
    subject_id = person_id,
    cohort_start_date = observation_period_start_date + 10,
    cohort_end_date = observation_period_start_date + 80
  )

exp2 <- obs |>
  transmute(
    cohort_definition_id = 2L,
    subject_id = person_id,
    cohort_start_date = observation_period_start_date + 150,
    cohort_end_date = observation_period_start_date + 220
  )

exp3 <- obs |>
  transmute(
    cohort_definition_id = 3L,
    subject_id = person_id,
    cohort_start_date = observation_period_start_date + 280,
    cohort_end_date = observation_period_start_date + 350
  )

df_exp <- bind_rows(exp1, exp2, exp3)

trt_set <- tibble(
  cohort_definition_id = 1:3,
  cohort_name = c("bortezomib", "lenalidomide", "daratumumab")
)

cdm <- insertTable(cdm = cdm, name = "mm_treatments", table = df_exp)
cdm$mm_treatments <- cdm$mm_treatments |> newCohortTable(cohortSetRef = trt_set)

# Generate Line of Therapy Subcohorts
cdm <- generateTherapyLineCohortSet(
  cdm = cdm,
  name = "mm_lot_cohorts",
  cohort = "mm_base",
  treatmentCohortName = "mm_treatments",
  gapEra = 60
)

# Inspect generated LOT cohorts
cdm$mm_lot_cohorts |>
  cohortCount()
```

---

# 2. Real-World Workflow with Concept Sets & DuckDB

In real-world OMOP CDM studies, base disease cohorts and treatment regimens are defined using concept sets and episode tables. The following pipeline (from `extras/generate_sankey_duckdb.R`) demonstrates the complete workflow on a DuckDB instance:

```r
library(DrugUtilisation)
library(CDMConnector)
library(omopgenerics)
library(OmopHelpers)
library(CohortConstructor)
library(duckdb)
library(htmlwidgets)

# Connect to DuckDB instance
con <- dbConnect(duckdb(), "test_data/omop_1959_1786811115.duckdb")
on.exit(dbDisconnect(con, shutdown = TRUE), add = TRUE)

cdm <- cdmFromCon(con, cdmSchema = "cdm", writeSchema = "main")

# 1. Build Base Multiple Myeloma Cohort (Concept Set 7343)
mmCodelist <- OmopHelpers::getCodelistFromConceptSet(
  conceptSetId = 7343,
  con = con,
  cdmSchema = "cdm"
)

cdm$base_mm <- CohortConstructor::conceptCohort(
  cdm = cdm,
  conceptSet = mmCodelist,
  name = "base_mm"
)

# 2. Extract Treatment Regimen Codelists (Concept Set IDs 7347-7378, 7483-7487, 11571-11573)
regimenIds <- c(7347:7358, 7360:7378, 7483, 7485:7487, 11571:11573)
rawCodelists <- lapply(regimenIds, function(id) {
  OmopHelpers::getCodelistFromConceptSet(conceptSetId = id, con = con, cdmSchema = "cdm")
})

mergedRaw <- unlist(rawCodelists, recursive = FALSE)
processedCodelists <- OmopHelpers::process_codelists(mergedRaw)
regimenCodelist <- omopgenerics::newCodelist(processedCodelists)

# 3. Generate Treatment Episode Cohort Set
cdm <- generateEpisodeCohortSet(
  cdm = cdm,
  name = "tx_episodes",
  conceptSet = regimenCodelist,
  episodeConceptId = 32531
)

# 4. Generate Line of Therapy Cohorts
cdm <- generateTherapyLineCohortSet(
  cdm = cdm,
  name = "lot_mm",
  cohort = "base_mm",
  treatmentCohortName = "tx_episodes",
  gapEra = 60
)

# 5. Render Interactive Plotly Sankey & Save Widget
sankey <- plotTherapyLineSankey(
  cdm$lot_mm,
  coverageThreshold = 0.80,
  maxLines = 4,
  minLinePatients = 5,
  includeEndOfFollowUp = TRUE
)

outputPath <- file.path(tempdir(), "mm_sankey_duckdb.html")
saveWidget(sankey, file = outputPath, selfcontained = TRUE)
```

---

# 3. Key Visualization Parameters

When calling `plotTherapyLineSankey()`, you can fine-tune diagram layout and aggregation thresholds:

- **`coverageThreshold = 0.80`**: At each line step, regimens representing the top $80\%$ cumulative volume are displayed individually, while long-tail regimens are grouped into `"Other Regimens"`.
- **`maxLines = 4`**: Maximum number of therapy line steps to display ($1\text{L}$ through $4\text{L}$).
- **`minLinePatients = 5`**: Minimum patient count required in a therapy line step to render the step column.
- **`includeEndOfFollowUp = TRUE`**: Terminal nodes indicating patient attrition or end of observation following Line $L$.

---

# Summary

- `generateTherapyLineCohortSet()` automates the construction of complex, multi-agent therapy lines in OMOP CDM datasets.
- `plotTherapyLineSankey()` produces clean, interactive Plotly visualizations suitable for observational research, clinical dashboards, and regulatory submissions.
